Saviors or Syndicates?: The Rise of Domestic Political Interventionism by Retired Military Leaders (Prospectus)

### 2020-10-23 code/prospectusv2.Rmd
commit

---
title: "Loyalty's Limits: The Impact of Postmodernism and Political Polarization on Military Professionalism  (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation seeks to explore the impact of increasing political polarization on the professional identity of military officers and the military institution. I develop a theory that argues that the nature of rising political polarization fundamentally challenges the conception of loyalty among members of the military profession, and that such a strain is problematic for an institution such as the military that holds the virtue of loyalty in high regard, and by statute possesses multiple, simultaneous loyalties: to the Constitution of the United States, to the President qua Commander-in-Chief, to Congress, to the American People, to one's fellow service members, and to one's self. Balancing these competing loyalties becomes more difficult as political polarization worsens, I argue, because of the fundamental and core values - not mere policy differences - that are at the root of many contemporary polarized political battles. Employing a mixed-methods approach that incorporates quantitative analysis and several historical case studies, this dissertation argues that the challenges to conceptions of loyalty among members of the military institution brought about by political polarization have the potential to impact normative concepts of military professionalism, which could be problematic not only for the character and identity of the military as an institution, but for American national security at large._   
geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

## Introduction and Research Question

In late May and early June 2020, racial tension and concerns about police brutality gripped the American public following the death of George Floyd, an unarmed black man, by police in Minneapolis. In many cities and communities across the nation, peaceful protests were followed in some cases by periods of rioting and looting. On June 1, after giving remarks at the Rose Garden, President Donald Trump walked to St. John's Church after security forces visibly displaced throngs of protesters in neighboring Lafayette Square. Flanked by members of his cabinet, to include the Secretary of Defense, and with the Chairman of the Joint Chiefs of Staff not far behind, the President held up a Bible at the church, stared into the cameras of the media, and, several minutes later, returned to the White House.

In the days and weeks that followed, a full-blown civil-military relations crisis ensued  [@feaver_military_2020]. The Secretary of Defense, Mark Esper, publicly disagreed with a threat made by President Trump to invoke the Insurrection Act of 1807 to use active-duty forces to quell domestic riots [@esper_secretary_2020], and General Mark Milley, the Chairman of the Joint Chiefs of Staff, told senior military officers graduating from the National Defense University that he "should not have been there (accompanying the President on the walk to St. John's Church)", as his presence "created a perception of the military involved in domestic politics" [@milley_official_2020].

Several retired military officers also weighed in, not only on the Trump administration's handling of specific events near the White House, but on a host of broader topics that dominated the popular dialogue in America after the death of George Floyd, to include the state of American racial relations. Upon paying close attention to what these retired military officers said, though, one can notice that there was in fact fairly substantial variation in both the substance and tone of their remarks. While all of them expressed concerns about the President potentially ordering active-duty military forces to confront what many Americans believed were peaceful protesters, and thus, violating Constitutional protections afforded by First Amendment [@brooks_dismay_2020; @dempsey_former_2020], some officers went much further in their criticisms. Indeed, some of these retired military officers accused the President of failing to possess the overall fitness to lead [@goldberg_james_2020], promote harmonious racial policies  [@allen_moment_2020], and demonstrate adequate leadership abilities [@mullen_i_2020]. In the months following George Floyd's death, moreover, several retired and currently-serving military officers have further made public arguments on a variety of topics that are directly related to President Donald Trump, including race in America [@askew_anti-racist_2020] and voter fraud as it relates to the upcoming Presidential Election [@nagl__2020].

While America has witnessed numerous incidents in its history in which tension between civilian and military elites has been present, the events of the Summer of 2020 between the Trump administration and some - though certainly not all - retired  military officers raises important questions, particularly I argue, because the context did not involve combat overseas.  These questions include: why did several retired military elites speak out so strongly against President Trump? Why did some of these elites confine their criticism to concerns over the Constitution, while others sounded more like representatives of the Democratic Party? Are these events that occurred during the Summer of 2020, and the military response generated in response to these events, an isolated chapter in American history, or is there something more disturbing underneath the surface regarding contemporary American civil-military relations?

This dissertation seeks to explore the impacts of rising domestic political polarization on the US military as an institution and profession. The primary argument espoused in this dissertation is that certain features of the polarized environment impact how military officers view and exercise the concept of loyalty, a claim that I argue holds significant ramifications for US national security. Using a mixed-methods approach, I develop a theory and then test it using a mixture of quantitative analysis and several qualitative case studies.  I conclude with discussing the implications for such an argument.

# Framing Within the Scholarly Literature

@@

## Civil Military Relations

Civil-military relations comprises an immense area of scholarship and interest, and deals with a correspondingly large number of questions and debates that have occupied the minds of political scientists, theorists, historians, ethicists, and sociologists for hundreds if not thousands of years. At its root, the study of civil-military relations examines what Peter Feaver calls the "problematique," or how to ensure that the military is strong enough to defeat threats to the polity without threatening the polity itself [@feaver_civil-military_1996; @feaver_civil-military_1999]. Feaver describes his "problematique" as "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1999, p. 214].

Civil-military relations involves a wide angle of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority, and a host of other massive and consequential issues impacting how militaries perform in war, their relationship to the societies they serve, and other important questions. In the context of the US, which has never experienced or come close to a coup (thankfully), the study of civil-military relations often focuses on the relationship between military and government elites, or what Risa Brooks calls "leadership at the state's apex" [@brooks_risa_integrating_2019, p. 380]. Indeed, it is at this level that vital decisions are made, including to go to war, how to prosecute war, how to achieve and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. Each of these questions, and many, many more, involve questions regarding the proper role of the military and its leaders relative to both society and, perhaps more important, the leaders of government. A number of scholars have examined this question on normative grounds, but none has been more lasting and consequential than the late Samuel Huntington, whose 1957 seminal work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for many scholarly works on civil-military relations. Although there have been many critiques of Huntington over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is absolutely essential to understand Huntington's concept of normative civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes the concept of what Huntington called "military professionalism" [@huntington_soldier_1957].

In _The Soldier and the State_, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims actually constrains the military in that the military can never become so alienated from society such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. The societal imperative, Huntington claimed, was a further function of two distinct features of the United States. The first was an "ideological constant," which referred to America's political ideology of liberalism [@huntington_soldier_1957, 143-162]. Huntington argued that liberalism as a political ideology stressed "individualism" [@huntington_soldier_1957, 90], and that American society was, since its founding, suspicious of standing military organizations and professional military forces. The second feature of the societal imperative, however, was the US Constitution, which Huntington claimed to be "conservative," and written by "men who feared concentrated political power and who provided for the widespread dispersion of that power among numerous governmental units" [@huntington_soldier_1957, 143]. The tension raised by these two imperatives - one functional stemming from external threat, the other societal stemming from the character of American political ideology and structure - needed to be solved, claimed Huntington. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society. Both goals - military security and civilian control - had to be obtained.

Figure 1, below, shows a visual diagram of Huntington's logic in a political science framework [@feaver_civil-military_1996]. As Peter Feaver points out in a 1996 article, Huntington did not use such language when developing and explaining his concept. According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that enables the achievement of these goals while also resolving the tension generated by the opposing functional and societal imperatives as the adoption of a type of civil-military relations framework which he calls, "objective civilian control" of the military [@huntington_soldier_1957, 83]. In such a model of "objective civilian control," Huntington envisioned civilians granting the military a high degree of professional autonomy, which he believed would simultaneously be "most conducive to the emergence of professional attitudes and bahvior among members of the officer corps" as well as the best way to minimize the military's political power by rendering the military "politically sterile and neutral" [@huntington_soldier_1957, 83-84], and thus preserving civilian supremacy.

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1: Samuel Huntington's Theory Of Civilan Control of the Military", subtitle="As Explained in The Soldier and The State, 1957")

Huntington's argument in support of "objective control" of the military stands in contra-distinction, as he explains, to an alternate model of "subjective control" in which civilian leaders limit the power of the military by treating it as any other civilian interest group [@huntington_soldier_1957, 80-83]. In fact, Huntington admits that the US Constitution actually envisioned and established a form of subjective vice objective scheme of civilian control of the military [@huntington_soldier_1957, 163-168]. What the framers did not see, nor could they, argues Huntington, was the rise of the military as a profession in response to ever-increasing and dangerous external threats. It was because of these pernicious threats, argues Huntington, which the US military had witnessed, confronted, and overcome during the 19th and 20th centuries in World War One, World War Two, and as Huntington writes, the Cold War, as well as the destructive capability of modern weapons and warfare in general that the military profession had emerged [@huntington_soldier_1957]. Huntington claims that unlike during the nation's founding, military officers constitute a profession, and thus, are unique and distinct from other occupations in several important ways: professions practice and develop a special body of knowledge and expertise; professions exercise a  "social responsibility" by providing a vital service to society; and professions have a degree of "corporateness" in that they develop and formalize standards by which members of the profession are certified to practice [@huntington_soldier_1957, 11-18].

Numerous critiques surrounding the Huntingtonian conception of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres, which Huntington to a certain degree certainly argued for, is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], charging that by its very nature, war requires the highest integration, as opposed to segregation, of political and military activities. Moreover, these critiques often point out, correctly in this author's opinion, that the conduct of war is more complicated than civilians simply telling the military the goals it wishes to achieve, and then allowing the military to determine how best to accomplish these goals. Rather, the development and implementation of strategy, especially during a time of war, requires an "iterative" dialogue between military leaders and civilians to ensure that effective policies are adopted [@rapp_civil-military_2015, 20].

Others argue that Huntington's conception of military professionalism is too narrow to be of practical use in the contemporary period in that such a model actually enables, rather than constrains, inappropriate political activity by military officers [@brooks_paradoxes_2020]. Risa Brooks charges that military officers who strongly identify with an ethic that stresses remaining apolitical at all costs, which she alleges follows from a Huntingtonian conception of military professionalism, may develop "blind spots such that they fail to recognize the political content or impact of their actions" [@brooks_paradoxes_2020, 17]. Whether military leaders recognize it or not, Brooks argues, military leaders comprise a profession that is inherently political, and such a reality ought to be acknowledged and embraced [@brooks_paradoxes_2020].

Regardless of whether and the degree to which one agrees or disagrees with Huntington's concept of military professionalism, a significant number of scholars note that a significant issue involves attempting to define where the line between appropriate political expression and inappropriate partistan activity should be drawn [@owens_military_2015; @golby_beyond_2015; @brooks_paradoxes_2020]. It is a difficult task for many reasons because, as Mackubin Thomas Owens points out, civil-military relations in the United States can be viewed as a "bargain" [@owens_us_2011] between the American people, the military, and the government that has to account for constantly-changing "political, social, technological, or geopolitical changes" [@owens_is_2016]. In other words, the ground is always moving for the actors involved in civil-military relations.

One of these recent changes, I argue, is increasing political polarization that has engulfed not only government elites, but the nation at large. Brooks (2020) points out three challenges posed by the contemporary period to the military's apolitical ethic, all of which are likely related to the growing problem of political polarization: rising acceptance among military officers of behaviors that contradict an apolitical ethos; increasing "pressures and opporunities" for military officers to express partisan sentiment; and increased politicization of the military by government elites [@brooks_paradoxes_2020, 21-23]. Indeed, as political polarization increases, the lines between partisanship and politics, as well as politics and everything not politics, have blurred, as so much of the literature on American politics and polarization affirms, promising that this challenge with respect to the military institution and its members will likely persist for some time.

@@

One potential area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity and character of the military. I am specifically interested in the question of if and how political polarization impacts the concept of loyalty in the military. This is an interesting research question because it brings together at least three relevant strands of literatures which, when combined, generates an interesting set of puzzles and questions. First, one of the most fundamental normative hallmarks of the military's professional nature, at least by Huntingtonian standards, is that it ought to be an apolitical and non-partisan institution. Second, one of the features of the US military is that it prizes and elevates the trait of loyalty among those who serve and, by statute as well as normative tradition, exercises simulataneous dual loyalties to its civilian heads, the President as the Commander in Chief, and Congress. And third, we know from the American politics literature that ideological polarization and partisan loyalty are connected in that those citizens who hold to stronger ideological convictions tend to be the most loyal partisans in terms of voting and other political behaviors [@abramowitz_great_2018, 112-116].

If these things are true, then a question is whether rising political polarization will generate instances in which military officers must choose between which objects of loyalty to ultimately elevate and side with in the event that there is conflict between and among these objects. Moreover, at the level of the military institution, a similar problem may also emerge: will the military institution have to choose between competing objects of loyalty, such as the President as the Commander in Chief, the Congress, and even its own professional ethic?^[Loyalty had to be a prime consideration in several prominent, recent cases involving the military and the government. Consider the case of now retired Army Lieutenant Colonel Alex Vindman, who testified before Congress against the US President, Donald Trump, regarding what Vindman believed to be behavior by the President that may have been illegal, or the case of Navy Captain Brett Crozier, who leaked a letter to the press after his pleas for help in battling the coranavirus aboard the ship he commanded, the USS Theodore Roosevelt, went unheeded for what he felt was too long. Certainly these instances involved individuals making choices about to whom or to what they would be most loyal to - the President, the Constitution, the ideals and obligations of the military, the operational mission, themselves as individual moral agents, and perhaps even more objects of loyalty.]

@@

This raises, then, an important question: does the military's essential purpose risk evolving or changing significantly during polarized times? While I believe that these potential developments are catching the attention of civil-military relations scholars, none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and in particular, the role of the military institution itself in facilitating these developments. Most scholars acknowledge the potentially damaging impacts of the politicization of the military [@golby_listening_2013, @brooks_paradoxes_2020] at the hands of civilian leaders, and I share these concerns. Fewer scholars, if any, are exploring the potential role of the military in politicizing itself, and the mechanisms through which these developments occur.

@@

# Initital Concerns and Timeline

>>>>

---
title: "Saviors or Syndicates?: The Rise of Domestic Political Interventionism by Retired Military Leaders (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation explores the changes in how a powerful block of military professionals - retired senior military leaders - intervene in domestic contemporary politics. I advance a theory which argues that changes in political activism by retired senior leaders is a result of two interacting factors: the level of domestic political polarization within a society, and the degree of prestige and credibility that military leaders enjoy within the society. To support my argument, I trace changes in levels and types of distinct forms of political activism, including authoring opinion-editorials in major newspapers, and participation in presidential endorsements. I further advance the theory by conducting a survey experiment that tests for differences between civilian and military responses after exposure to situations that evoke one's partisan or ideological identity. I ultimately argue that the increasing political activism of retired military leaders is not merely a response to the hyper-partisanship present in the current political atmosphere, but rather the logical outworking of trends that have been active for some time in which both the ideological differences between Americans and the prestige of the military have increased. I conclude by describing the effect of the implications of growing politicla activism by senior military leaders on the core beliefs of the military regarding its very own essence and role in society._

geometry: margin = 1.15in              # margin widths
fontsize: 12pt                         # text size
bibliography: researchpaper-ps904.bib # bibliography file
biblio-style: authoryear               # citation style
urlcolor: blue                         # links to external URLs
citecolor: black                       # links to bibliography
linkcolor: magenta                     # internal figures, footnotes, etc.
indent: true                          # paragraph format
                                       # true for indented paragraphs

@@

# Introduction and Research Question

Across the world and in almost nearly every country, discussions regarding the participation of the military in politics are nothing new. Indeed, few topics evoke such strong opinions and emotions than whether and to what degree the military, both as a corporate body and as individual military officers, should be political vice remain separate from certain aspects of and activities associated with politics. These normative questions rightly form a significant portion of the study of civil-military relations, and intellectuals, practitioners, and pundits have frequently weighed in on these topics over the years.

Thankfully, within the United States, unlike many other countries across the world and throughout history, there has never been a coup, nor a serious attempt made by the military to dislodge political control from their duly-constituted elected leaders and to hold it themselves. However, as almost every civil-military relations scholar points out, the absence of coups does not mean that the relationship between civilian and military leaders is healthy, nor absent of conflict altogether. Indeed, a cursory examination of the relationships between Lincoln and MacClellan, Truman and MacArthur, and Obama and McChrystal reveals that disagreement between those who serve at the highest levels of civilian government and military organizations is neither rare nor understudied.

What has changed, however, is the degree to which a very specific subset of the American military institution - retired senior military officials - have participated in the domestic political process post retirement. Indeed, whereas General Omar Bradley, one of the military's five star generals and a hero of World War Two, famously remarked in 1959, "I am convinced that the best service a retired general can perform is to turn in his tongue along with his suit and mothball his opinions," [@times_bradley_1959], retired generals and admirals today appear frequently in the media, endorse Presidential candidates at higher frequency [@griffiths_not_2019], and frequently serve in government after their retirement from military service [@brooks_dont_2016]. To provide one such example, following protests across the United States in the summer of 2020 in which President Donald Trump nearly ordered the use of active duty military forces in American cities, former Chairman of the Joint Chiefs of Staff Admiral (Ret) Michael Mullen was one of at least ten four star retired officers who publicly aired concern about the President's potential use of force. ^[Other officers included General (Ret) James Mattis, General (Ret) Mark Kelly, Admiral (Ret) William McRaven, General (Ret) Martin Dempsey, General (Ret) Colin Powell, General (Ret) Michael Hayden, General (Ret) David Petraeus, General (Ret) John Allen, and General (Ret) Vincent Brooks.]. During an interview in the days which followed, when asked by NPR reporter why he spoke out, Mullen stated, "we were getting to a point where the country was in jeopardy, which is why I spoke out" [@isaacson_ret_2020; for further analysis, see also @brooks_let_2020].

Mullen's words are incredible from a number of important perspectives, and raise important normative questions regarding the military's role in a democratic society [@brooks_let_2020] and the conditions under which the military can and should express dissent to its civilian bosses [@cohn_what_2019]. Few, if any, scholars have attempted to systematically examine and explain the rise of political activism by retired military officers over time. Such an effort would better illumine the understanding of scholars, as well as the American military and society, of why a powerful group of what were once non-political actors have, at least in some cases, suddenly become one.  More important, a systematic study of this trend would contribute to a much-needed discussion regarding the role of the military institution and its relationship to the government and American society in a contentious political age.

The central argument made in this dissertation is that increasing political activism at the hands of retired military officers is the result of two longstanding and interconnected trends: rising political polarization within the domestic political sphere, and rising prestige and credibility of military institutions afforded by American society. The interaction of these trends explains the remarkable increase in political activity by retired senior military leaders, and portends a shift in the perceived role that these senior military leaders fill vis-a-vis civilian authority - a shift away from professional and obedient servants and towards "guardians of democracy" who pick and choose when and how to intervene in American politics. Such a shift builds on previous work regarding normative "role beliefs" [@fitch_armed_1998] of the military institution, and has significant ramifications on the organizational culture of the military, its relationship with government and society, and American national security.

# The Field of Civil-Military Relations

Civil-military relations involve a wide host of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority between military and civilian leaders, and a number of other consequential issues. At its root, the study of civil-military relations examines what Peter Feaver calls the "problematique," or how to ensure that the military is strong enough to defeat threats to the polity without threatening the polity itself [@feaver_civil-military_1996; @feaver_civil-military_1999]. Feaver describes this "problematique" as "vexing" and a great "paradox" in that the military must be "strong enough to prevail in war," yet submissive to authority such that it cannot and will not "destroy or prey on the society it is intended to protect" [@feaver_civil-military_1999, p. 214]. Within the context of the US, the study of civil-military relations often focuses on the relationship between military and government elites, or what Risa Brooks calls "leadership at the state's apex" [@brooks_risa_integrating_2019, p. 380]. Indeed, it is at this level that vital decisions are made, including to go to war, how to prosecute war, how to achieve and implement effective strategy, how to prepare for the next conflict, and how large the military and its budget should be. A number of scholars have offered solutions to Feaver's "problematique," but none has been more lasting and consequential than the late Samuel Huntington, whose 1957 seminal work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, often serves as the point of departure for many scholarly works on civil-military relations. Although there have been many critiques of Huntington over the years [for example, see @janowitz_professional_1960; @finer_man_1962; @cohen_supreme_2003; @feaver_armed_2003; @brooks_paradoxes_2020], it is absolutely essential to understand Huntington's concept of normative civil-military relations, and what lies at the heart of it - a theory of civilian control of the military that maximizes the concept of what Huntington called "military professionalism" [@huntington_soldier_1957].

In _The Soldier and the State_, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims actually constrains the military in that the military can never become so alienated from society such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. The societal imperative, Huntington claimed, was a further function of two distinct features of the United States. The first was an "ideological constant," which referred to America's political ideology of liberalism [@huntington_soldier_1957, 143-162]. Huntington argued that liberalism as a political ideology stressed "individualism" [@huntington_soldier_1957, 90], and that American society was, since its founding, suspicious of standing military organizations and professional military forces. The second feature of the societal imperative, however, was the US Constitution, which Huntington claimed to be "conservative," and written by "men who feared concentrated political power and who provided for the widespread dispersion of that power among numerous governmental units" [@huntington_soldier_1957, 143]. Huntington claimed that the tension generated by these two imperatives - one functional stemming from external threats, the other societal stemming from the character of American political ideology and structure - needed to be solved. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society. Both goals - military security and civilian control - had to be obtained.

Figure 1, below, shows a visual diagram of Huntington's logic in a political science framework [@feaver_civil-military_1996]. As Peter Feaver points out in a 1996 article, Huntington did not use such language when developing and explaining his concept. According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that enables the achievement of these goals while also resolving the tension generated by the opposing functional and societal imperatives as the adoption of a type of civil-military relations framework which he calls, "objective civilian control" of the military [@huntington_soldier_1957, 83]. In such a model of "objective civilian control," Huntington envisioned civilians granting the military a high degree of professional autonomy, which he believed would simultaneously be "most conducive to the emergence of professional attitudes and behavior among members of the officer corps" as well as the best way to minimize the military's political power by rendering the military "politically sterile and neutral" [@huntington_soldier_1957, 83-84], and thus preserving civilian supremacy.

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1: Samuel Huntington's Theory Of Civilian Control of the Military", subtitle="As Explained in The Soldier and The State, 1957")

Huntington's argument in support of "objective control" of the military stands in contra-distinction, as he explains, to an alternate model of "subjective control" in which civilian leaders limit the power of the military by treating it as any other civilian interest group [@huntington_soldier_1957, 80-83]. In fact, Huntington admits that the US Constitution actually envisioned and established a form of subjective vice objective scheme of civilian control of the military [@huntington_soldier_1957, 163-168]. What the framers did not see, nor could they, argues Huntington, was the rise of the military as a profession in response to ever-increasing and dangerous external threats. It was because of these pernicious threats, argues Huntington, which the US military had witnessed, confronted, and overcome during the 19th and 20th centuries in World War One, World War Two, and as Huntington writes, the Cold War, as well as the destructive capability of modern weapons and warfare in general, that the military profession had emerged [@huntington_soldier_1957]. Huntington claims that Cold-War military officers, unlike the corporate body of officers that existed at the nation's founding, constitute a profession and thus, are part of a unique and distinct organization that differs from other occupations: professions practice and develop a special body of knowledge and expertise; professions exercise a  "social responsibility" by providing a vital service to society; and professions have a degree of "corporateness" in that they develop and formalize standards by which members of the profession are certified to practice [@huntington_soldier_1957, 11-18].

Numerous critiques surrounding the Huntingtonian conception of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres, which Huntington to a certain degree certainly argued for, is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], and thus point out that by its very nature, war requires the highest integration - not segregation - of political and military activities. Moreover, these critiques often point out, correctly in this author's opinion, that the conduct of war is more complicated than civilians simply telling the military the goals it wishes to achieve, and then allowing the military to determine how best to accomplish these goals. Rather, the development and implementation of strategy, especially during a time of war, requires an "iterative" dialogue between military leaders and civilians to ensure that effective policies are adopted [@rapp_civil-military_2015, 20].

Others argue that Huntington's conception of military professionalism is too narrow to be of practical use in the contemporary period in that such a model actually enables, rather than constrains, inappropriate political activity by military officers [@brooks_paradoxes_2020]. Risa Brooks charges that military officers who strongly identify with an ethic that stresses remaining apolitical at all costs, which she alleges follows from a Huntingtonian conception of military professionalism, may develop "blind spots such that they fail to recognize the political content or impact of their actions" [@brooks_paradoxes_2020, 17]. Whether military leaders recognize it or not, Brooks argues, military leaders comprise a profession that is inherently political, and such a reality ought to be acknowledged and embraced [@brooks_paradoxes_2020]. Regardless of whether and the degree to which one agrees or disagrees with Huntington's concept of military professionalism, a significant number of scholars note that a significant issue involves attempting to define where the line between appropriate political expression and inappropriate partisan activity should be drawn [@owens_military_2015; @golby_beyond_2015; @brooks_paradoxes_2020]. It is a difficult task for many reasons because, as Mackubin Thomas Owens points out, civil-military relations in the United States can be viewed as a "bargain" [@owens_us_2011] between the American people, the military, and the government that has to account for constantly-changing "political, social, technological, or geopolitical changes" [@owens_is_2016]. In other words, the ground is always moving for the actors involved in civil-military relations.

## The Dependent Variable: Military Intervention in Politics and Military "Role Beliefs"

Given the normative model espoused by Huntington and the critiques his model has been subjected to, it is important to understand what frameworks exist for understanding why the military would intervene in politics. Numerous scholars, mainly writing from a comparative perspective (although not exclusively), have enriched our understanding of what possibly motivates the military to interject itself into the political sphere. To be clear, these scholars are focusing mainly on active-duty military organizations, and not on the specific block of retired military officers, as I plan to examine, but the motivations and mechanisms for intervention that they put forward enrich our understanding of the phenomenon of political intervention by the military.

The British political scientist, S.E. Finer, argues that the military's intervention in politics is a function of the opportunity and desire to do so [@finer_man_1962]. Finer identifies several factors which should normatively inhibit the military from intervening in politics. These factors are military professionalism, the principle of civilian supremacy, and fear that the military would fail to remain a cohesive fighting force should it fracture internally upon intervening in political matters [@finer_man_1962]. Finer critiques the concept of Huntingtonian military professionalism somewhat harshly, arguing that although military professionalism should act to focus the military on itself and thus not on political activities, those who claim to be professional might and have used the concept of professionalism as an excuse to not obey civilian authorities [@finer_man_1962]. Finer is clear that the principle of civilian control - not military professionalism - is the true and only bulwark that can keep the military out of politics.

Describing the various motives that militaries might have for intervening in politics, Finer goes on to describe at least three. He first lists the motive of the "manifest destiny of the soldiers," in which the military acts as "saviors" of the country [@finer_man_1962, 28-29]. This motive, Finer claims, arises because of the confluence of the military's unique mission, sense of power, and claim of being "outside party politics" [@finer_man_1962, 29]. A second motive that militaries might possess which disposes them to intervene in politics is what Finer calls the "unique identification with the national interest" [@finer_man_1962, 30]. Acting from such a motive, Finer conceptualizes the military acting as "custodians" of the "national interest" [@finer_man_1962, 30]. Finer acknowledges that "custodianship" can take on different forms in different contexts, but points out that some militaries might view "custodianship" as a "duty to arbitrate or veto...to exercise it when some convulsion or decision of the civil autorities seems to them [the military] to threaten what they think are the permanent interests of the nation" [@finer_man_1962, 31]. The final motive Finer spends the most time on, calling it "the sectional interest," which refers to a more narrow subset of interests than the "national interest," and includes a wide ranging set of interests based on social class, regional affiliation, personal career motivations, or organizational interests [@finer_man_1962, 34-53].

Ultimately, Finer develops a typology of four levels, or degress of, military intervention. The first he calls "influence," and argues that at this level of intervention, the military makes appeals "to reason or emotion" in an attempt to persuade the civilian government [@finer_man_1962, 77]. Finer adds that intervening at this level is, in his mind at least, not problematic and entirely "consistent with the supremacy of the civil power" [@finer_man_1962, 77]. The next level of intervention is "pressures or blackmail," describing efforts by the military to "convince the civil power by threat of some sanction" [@finer_man_1962, 77]. Finer adds that military intervention via this level is "on the margin," referring to his belief that while some cases may be unproblematic, many other forms, such as the "threat to stage a coup," would be normatively wrong [@finer_man_1962, 77]. Finer points out a third level of intervention, that of "displacement," referring to the "removal" of government officials by the military as a result of "violence or the threat of violence" [@finer_man_1962, 78]. The fourth level, which is basically displacement on a wider scale, Finer calls "supplantment," which "sweeps away the civilian regime and establishes the military in its place" [@finer_man_1962, 78]. Finer goes on to explain the level of intervention as a function of the level of "political culture" that is prevalent in a state and society  [do I need more here? See Finer, pp. 79 and following]

Another helpful conceptualization of military intervention in politics is the idea of "role beliefs," a concept articulated by the political scientist, J. Samuel Fitch, in his book, _The Armed Forces and Democracy in Latin America._

A third, and equally rich, study of the degrees of military intervention in politics is available from Brian Taylor's book, _Politics and the Russian Army: Civil-Military Relations, 1689-2000.

@@

@@

One potential area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity and character of the military. I am specifically interested in the question of if and how political polarization impacts the concept of loyalty in the military. This is an interesting research question because it brings together at least three relevant strands of literatures which, when combined, generates an interesting set of puzzles and questions. First, one of the most fundamental normative hallmarks of the military's professional nature, at least by Huntingtonian standards, is that it ought to be an apolitical and non-partisan institution. Second, one of the features of the US military is that it prizes and elevates the trait of loyalty among those who serve and, by statute as well as normative tradition, exercises simultaneous dual loyalties to its civilian heads, the President as the Commander in Chief, and Congress. And third, we know from the American politics literature that ideological polarization and partisan loyalty are connected in that those citizens who hold to stronger ideological convictions tend to be the most loyal partisans in terms of voting and other political behaviors [@abramowitz_great_2018, 112-116].

If these things are true, then a question is whether rising political polarization will generate instances in which military officers must choose between which objects of loyalty to ultimately elevate and side with in the event that there is conflict between and among these objects. Moreover, at the level of the military institution, a similar problem may also emerge: will the military institution have to choose between competing objects of loyalty, such as the President as the Commander in Chief, the Congress, and even its own professional ethic?^[Loyalty had to be a prime consideration in several prominent, recent cases involving the military and the government. Consider the case of now retired Army Lieutenant Colonel Alex Vindman, who testified before Congress against the US President, Donald Trump, regarding what Vindman believed to be behavior by the President that may have been illegal, or the case of Navy Captain Brett Crozier, who leaked a letter to the press after his pleas for help in battling the coronavirus aboard the ship he commanded, the USS Theodore Roosevelt, went unheeded for what he felt was too long. Certainly these instances involved individuals making choices about to whom or to what they would be most loyal to - the President, the Constitution, the ideals and obligations of the military, the operational mission, themselves as individual moral agents, and perhaps even more objects of loyalty.]

@@

This raises, then, an important question: does the military's essential purpose risk evolving or changing significantly during polarized times? While I believe that these potential developments are catching the attention of civil-military relations scholars, none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and in particular, the role of the military institution itself in facilitating these developments. Most scholars acknowledge the potentially damaging impacts of the politicization of the military [@golby_listening_2013; @brooks_paradoxes_2020] at the hands of civilian leaders, and I share these concerns. Fewer scholars, if any, are exploring the potential role of the military in politicizing itself, and the mechanisms through which these developments occur.

@@

# Initial Concerns and Timeline
